# Reproducible Research: Peer Assessment 1
=========================================
This version of the PA1_template was written by **Jacob Spangler**.

## Download and unzip the dataset
```{r Download and Unzip Dataset, echo=TRUE}
#The raw data should be stored in Github with this script. If it isn't then this portion will 
#download the file and then unzip it.
if(!file.exists("activity.zip")){
  #Identify the location of the datafile
  fileURL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip"
  download.file(fileURL, "activity.zip", mode="wb")
  dateDownloaded <- date()
  print(dateDownloaded)
  
  #Unzip the data
  unzip("activity.zip")
}else{
  unzip("activity.zip")
}
```

## Loading and preprocessing the data
```{r Loading and preprocessing data, echo=TRUE}
#Read in the data
Activity_Data <- read.csv("activity.csv")

#Format the Dates
Activity_Data$date <- as.Date(Activity_Data$date, "%Y-%m-%d")

#Format the interval to 24 hr time clock
Activity_Data$time <- ifelse(nchar(Activity_Data$interval) == 1,
                             paste(000, Activity_Data$interval, sep=""),
                             Activity_Data$interval)
Activity_Data$time <- ifelse(nchar(Activity_Data$time) == 2,
                             paste(00, Activity_Data$time, sep=""),
                             Activity_Data$time)
Activity_Data$time <- ifelse(nchar(Activity_Data$time) == 3,
                             paste(0, Activity_Data$time, sep=""),
                             Activity_Data$time)

#Complete the date and interval formatting
Activity_Data$Date_Time <- as.POSIXct(paste(Activity_Data$date, Activity_Data$time),
                                      format="%Y-%m-%d %H%M")
```

## What is mean total number of steps taken per day?
```{r fig.width=8, fig.height=8, echo=TRUE}
library(ggplot2)

ggplot(Activity_Data) +
  geom_histogram(aes(x=date, y=steps),  stat="identity") +
  ylab("Steps") +
  xlab("Date") +
  ggtitle("Steps taken per day")
```

### What is the average number of steps per day?
```{r Mean steps per day, echo=TRUE}
tapply(Activity_Data$steps, Activity_Data$date, mean, na.omit=TRUE)
```

### What is the median number of steps per day?
```{r Median steps per day, echo=TRUE}
tapply(Activity_Data$steps, Activity_Data$date, median, na.rm=TRUE)
```

## What is the average daily activity pattern?
```{r fig.width=8, fig.height=8, echo=TRUE}
Removed_NAs <- na.omit(Activity_Data)
Avg_By_Interval <- as.data.frame(tapply(Removed_NAs$steps, Removed_NAs$time, mean))
colnames(Avg_By_Interval) <- "Average"
Avg_By_Interval$Interval <- rownames(Avg_By_Interval)

plot(Avg_By_Interval$Interval, Avg_By_Interval$Average, type="l",
     ylab="Average Steps", xlab="Time Interval", main="Steps per 5 minute interval")

High_Int <- Avg_By_Interval$Interval[Avg_By_Interval$Average==max(Avg_By_Interval$Average)]
```
The 5-minute interval with the highest average number of steps is **`r High_Int`**.

## Imputing missing values
```{r Count of Missing Values, echo=TRUE}
Missing_value_Count <- length(Activity_Data$steps[is.na(Activity_Data$steps)])
```
The number of missing values in the activities dataset is **`r Missing_value_Count`**.

This code will fill in the missing values using the average number of steps for the given interval
```{r Impute Missing Data, echo=TRUE}
Activity_Data$Imputed_Data <- Activity_Data$steps

for(i in 1:nrow(Activity_Data)){
  if(is.na(Activity_Data$Imputed_Data[i])){
    Values_Interval <- Activity_Data$time[i]
    Activity_Data$Imputed_Data[i] <- as.numeric(Avg_By_Interval$Average[Avg_By_Interval$Interval==Values_Interval])
  }else{
    Activity_Data$Imputed_Data[i] <- Activity_Data$Imputed_Data[i]
  }
}

```

This will create a histogram using the imputed values.
```{r fig.width=8, fig.height=8, echo=TRUE}
library(ggplot2)

ggplot(Activity_Data) +
  geom_histogram(aes(x=date, y=Imputed_Data),  stat="identity") +
  ylab("Steps") +
  xlab("Date") +
  ggtitle("Steps taken per day (with Imputed Values)")
```

### What is the average number of steps per day (with Imputed Values)?
```{r Mean steps per day Imputed, echo=TRUE}
tapply(Activity_Data$Imputed_Data, Activity_Data$date, mean, na.omit=TRUE)
```

### What is the median number of steps per day (with Imputed Values)?
```{r Median steps per day Imputed, echo=TRUE}
tapply(Activity_Data$Imputed_Data, Activity_Data$date, median, na.rm=TRUE)
```

## Are there differences in activity patterns between weekdays and weekends?
```{r fig.width=8, fig.height=8, echo=TRUE}

Activity_Data$Day_of_Week <- weekdays(Activity_Data$date)
Activity_Data$Category <- "Weekday"
Activity_Data$Category[Activity_Data$Day_of_Week=="Sunday"] <- "Weekend"
Activity_Data$Category[Activity_Data$Day_of_Week=="Saturday"] <- "Weekend"
Activity_Data$Category <- factor(Activity_Data$Category, levels=c("Weekday", "Weekend"))

ggplot(Activity_Data, aes(x=interval, y=Imputed_Data)) +
  geom_line() +
  facet_grid(Category~.) +
  xlab("Time Interval") +
  ylab("Average Steps") +
  ggtitle("Steps per 5 minute interval")
```

